---
title: "Activity Monitor"
author: "Bernard Fraenkel"
date: "September 19, 2015"
output: html_document
---
```{r setoptions, echo=FALSE}
knitr::opts_chunk$set(echo=TRUE,warning=FALSE)
```

# Introduction

## Loading libraries and raw data 

```{r, results=FALSE, message=FALSE}
# Code for Project 1 of Coursera Reproducible Research Course
# 9/16/2015

library(ggplot2)
library(gridExtra)
library(plyr) # Has to be before dplyr
library(dplyr)

# 1. Load the data
activity <- read.csv("activity.csv")
message(paste0("#Records in activity: "), nrow(activity))
```

## Processing the raw data into a format suitable for your analysis

The first remove records that have NA values
```{r, message=FALSE}
# Filter out the NAs
activity_clean <- activity[complete.cases(activity),]
message(paste0("#Records in activity_clean: "), nrow(activity_clean))
```
## Mean total number of steps taken per day

``` {r, message=TRUE}
# Calculate the total number of steps taken per day
daily_steps <- ddply(activity_clean, .(date), summarize, sum_steps=sum(steps))
av_daily_steps <- mean(daily_steps$sum_steps)
message(paste0("Mean total number of steps taken per day: "), round(av_daily_steps,0))
```

The plot below shows the total number of steps taken each day

```{r}
g <- ggplot(daily_steps,aes(date, sum_steps))
g <- g + geom_point() + ggtitle("Total number of steps taken per day")
g
```

## Histogram of the total number of steps taken each day
```{r}
qplot(sum_steps, data=daily_steps, geom="histogram", main="Histogram of total number of steps per day")
```

## Mean and median of the total number of steps taken per day
``` {r, message=TRUE}
mean_steps <- mean(daily_steps$sum_steps)
# mean_steps <- round(mean_steps,9)
median_steps <- median(daily_steps$sum_steps)
# median_steps <- round(median_steps,0)
message(paste0("daily average number of steps:", round(mean_steps,2)))
message(paste0("median of steps taken per day:", round(median_steps,2)))
```
  


## Average daily activity pattern

This is a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
interval_steps <- ddply(activity_clean, .(interval), summarize, sum_steps=sum(steps))
nb_days <- nrow(daily_steps)  # Number of days for which we have observations
# compute the average number of steps per interval
av_steps <- mutate(interval_steps, sum_steps = sum_steps/nb_days)
colnames(av_steps) <- c("interval", "avrg_steps")
g <- ggplot(av_steps, aes(interval, avrg_steps))
g <- g+geom_line(size=1,colour="blue")  + ggtitle("Daily average number of steps per interval")
g
```

## Interval with maximum number of steps

The  5-minute interval, on average across all the days in the dataset, which contains the maximum number of steps is 
```{r}
av_steps_ordered <- av_steps[with(av_steps, order(-avrg_steps)),]
message(paste0("Interval with maximum number of steps: ", av_steps_ordered[1,"interval"]))
```

The time of *maximum activity is thus around 8:30 ins the morning*  
The plot also shows minor peaks at the typical times of activity: 12pm, 3:30pm and 7pm

# Imputing missing values

## Total number of missing values in the dataset
```{r}
activity_NA <- filter(activity, is.na(steps))
na_rows <- nrow(activity_NA)
message(paste0("#rows with missing data: "), na_rows, " - out of a total of: ", nrow(activity))
```


## Filling in all of the missing values in the dataset

The code below fills in missing data for a given interval with the overall average value for that interval

```{r}

# Replace NA values with the average values obtained from records with non-missing values: av_steps
# Create a new dataset that is equal to the original dataset but with the missing data filled in.
# Round the averages 
# activity_full <- mutate(activity, steps = ifelse(is.na(steps), round(av_steps[av_steps$interval == interval, "avrg_steps"],0), steps))
activity_full <- mutate(activity, steps = ifelse(is.na(steps), round(av_steps[match(interval,av_steps$interval), "avrg_steps"],0), steps))
if(any(is.na(activity_full))) stop ("still some NAs")
```


## Histogram of the total number of steps taken each day
```{r}
daily_steps_full <- ddply(activity_full, .(date), summarize, sum_steps=sum(steps))
qplot(sum_steps, data=daily_steps_full, geom="histogram",  main="Histogram of total number of steps per day")

# Calculate and report the mean and median total number of steps taken per day.
mean_steps_full <- mean(daily_steps_full$sum_steps)
median_steps_full <- median(daily_steps_full$sum_steps)

message(paste0("daily average number of steps - after NA replacement:", round(mean_steps_full,2)))
message(paste0("median of steps taken per day  - after NA replacement:", round(median_steps_full,2)))
```

These values differ from the estimates from the first part of the assignment   


Imputing missing data on the estimates of the total daily number of steps?

## Differences in activity patterns between weekdays and weekends
Note that we use the dataset with the filled-in missing values for this part  
```{r}
# Function to convert weekday names to weekday/weekend
day_vs_end <- function(date_str) {
	wkday <- weekdays(as.Date(date_str))
	ifelse(wkday == "Saturday" | wkday == "Sunday", "weekend", "weekday")
}
activity_final <- mutate(activity_full, day_vs_end = day_vs_end(date))
activity_final <- mutate(activity_final, day_vs_end = as.factor(day_vs_end))
# Split data set into 2
activity_weekday <- activity_final[activity_final$day_vs_end == "weekday", c("interval", "steps")]
activity_weekend <- activity_final[activity_final$day_vs_end == "weekend", c("interval", "steps")]
# interval_steps_final <- ddply(activity_final, .(interval), summarize, sum_steps=sum(steps))


# Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis)
# and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
nb_wkdy <- nrow(activity_weekday)/288  # 288 5-min intervals per day
nb_wknd <- nrow(activity_weekend)/288
interval_steps_weekday <- ddply(activity_weekday, .(interval), summarize, av_steps=sum(steps)/nb_wkdy)
interval_steps_weekend <- ddply(activity_weekend, .(interval), summarize, av_steps=sum(steps)/nb_wknd)
message(paste0("Average number of steps per 5-min on weekdays: ", round(sum(interval_steps_weekday$av_steps),2)))
message(paste0("Average number of steps  per 5-min on weekend days: ", round(sum(interval_steps_weekend$av_steps),2)))
```

This shows that the subject is about _20% more active during weekends_

The 2 plots below compare activity during weekdays versus weekends
- During weekdays, activity is  concentrated mostly around 8:30am - one can assume this is time reserved for working out  
- During weekdends, overall activity starts much later (9am vs 5am) compared to weekdays - it is also roughly evenly spread out during the day (vs weekdays when activity is concentrated at "work-out" times)

```{r}
g <- ggplot(interval_steps_weekday,aes(interval, av_steps))
g <- g+geom_line(size=1,colour="blue") + ggtitle("Daily activity during weekdays") + ylim(0,250)
q <- ggplot(interval_steps_weekend,aes(interval, av_steps))
q <- q+geom_line(size=1,colour="red") + ggtitle("Daily activity during weekend") + ylim(0,250)
grid.arrange(g, q, nrow=2)
```
